Embedded Domain Specific Languages inside Functional Languages

### 2014-03-02 paper.tex
Adding QC example

An embedded DSL (EDSL) is a language inside a language.
An EDSL implemented as a library in a host language
that has the look, feel and semantics of its own language,
customized to a specific problem domain.
Haskell~\cite{Haskell98Book}, our functional language of choice, is a great host for EDSLs
because of flexible overloading, a powerful type system, and lazy semantics facilitating this.
The functional programming community has taken  the ideas
of EDSLs, which considerably lowered the costs of developing
and maintaining a DSL, and developed a number of DSL 
which provide higher-level interfaces and abstractions for well-understood systems.

@@

\begin{Code}
Prelude Test.QuickCheck> quickCheck prop_reverse 
+++ OK, passed 100 tests.
\end{Code}

>>>>

An embedded DSL (EDSL) is a language inside a language.
An EDSL implemented as a library in a host language
that has the look, feel and semantics of its own language,
customized to a specific problem domain.
Haskell~\cite{Haskell98Book}, our functional language of choice, is a great host for EDSLs
because of flexible overloading, a powerful type system, and lazy semantics facilitating this.

As an example of Haskell, consider reversing a list in Haskell
\begin{Code}
reverse :: [a] -> [a]
reverse []     = []
reverse (x:xs) = reverse xs ++ [x]
\end{Code}

This is the complete defintion of reverse: reverse reverses any type of list (the first line);
the reverse of an empty list, written \verb|[]|, is an empty list;
and the reverse of an non-empty list is the reverse of the tail of a linked-list,
appended to the front of the list.

Using this consise syntax, Haskell specifically and the functional programming
community in general have taken the ideas
of EDSLs, which considerably lowered the costs of developing
and maintaining a DSL, and developed many, many of DSL 
which provide higher-level interfaces and abstractions for well-understood systems.

@@

\verb|prop_reverse| is itself a regular Haskell function,
that takes a list of Int, and returns a boolean, based
on the validity of what is being proposed, specifically
that is a two reverses cancel each other out. 
Here is the neat part -- \verb|prop_reverse| is {\bf also\/} a
domain specific statement, and as such can also be considered
a sub-language inside of Haskell itself.  This style
of using function, in this case function with names
prefixed with \verb|prop_|, taking a number of typed
arguments, and returning a conditional. The property
written in Haskell is also an Embeeded Domain specific
language for properties.

We can run our EDSL, using a function called \verb|quickCheck|:
\begin{Code}
Prelude Test.QuickCheck> quickCheck prop_reverse 
+++ OK, passed 100 tests.
\end{Code}
By running \verb|quickCheck| with our explicit and specific property,
we execute our ESDL inside Haskell. The quickCheck function
generates 100 test cases for our property, and exectutes
them on the fly. If they all hold, then the system prints
a message reflecting this.

In turns out that this sort of mini-language is really useful in practice.
